\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}

\input{zMC.tex}
\input{zMCgraphe.tex}

\zzpackages[english]
\zzhyperref
\zzmarges


\zzheader{PGM Project}{}{\today}{De Lara, Tilquin, Vidal}{}{\arabic{page}/\pageref{lastpage}}

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                MACRO LOCALES                                 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%

% Divergence de Kullback-Leibler
\newcommand{\Kl}[3][]{\mathrm K_{#1}\!\zp{#2\:\|\:#3}}
\newcommand{\zZ}[2]{\mathrm #1\!\zp{#2}}
\newcommand{\zD}{\mathcal}
\newcommand{\Ng}[2]{\mathcal{N}\zp{#1,\:#2}}

\tikzset{
  zplot/.style={opacity=.8}
}

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                DEBUT DOCUMENT                                %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%

\begin{document} 

\def\foo#1#2{\vbox{\hbox to 2cm{\hss\scshape #1\hss}\hbox to 2cm{\hss\footnotesize#2\hss}}\hss}

\begingroup\centering
{\bfseries \huge Independent Component Analysis}\par\vspace{.3cm}
{\bfseries \Large PGM Project}\par\vspace{.7cm}
\hbox to \textwidth{\hss
\foo{Nathan de Lara}{École polytechnique}
\foo{Florian Tilquin}{ENS Cachan}
\foo{Vincent Vidal}{ENS Ulm}
}\par\vspace{.8cm}
\today\zal\vspace{.3cm}\zal
\zligne\endgroup

\tableofcontents

\vspace{1cm}\zligne


\begin{center}\textsc{Abstract}\end{center}
This paper is dedicated to the study of Independent Component Analysis. We intent to implement, apply and compare several algorithms while being presenting some theoretical aspects such as the link between the likelihood maximisation and the mutual information.

%% ------- -- -- -  -                                      -  - -- -- --------%%
%---- -- -  -                     Introduction                     -  - -- ----%
%%-------- -- -- -  -                                      -  - -- -- ------- %%


\section{Problem statement}
\subsection{Introduction}

The general Independent Component Analysis problem can be formalised this way:
Suppose we have some random variables $x\in\zR^p$ which correspond to
a mix of some primitive sources $s\in\zR^n$. The aim is to extract
from $x$ every source $s_i$. To do so, we will suppose here that:
\zitemize{
\item[--] the sources are independents.
\item[--] the mix is linear and instantaneous
\item[--] at most one source has a Gaussian distribution.
}
We define:
\begin{equation}
  x = A s \ \ \mbox{and} \ \ y = W x,
\end{equation}
where $A$ is the mixing matrix, $W$ the separation matrix and $y$ the
estimation of the sources. The goal is then to find a matrix $W$ that
maximise a certain measure of independence of $y$.

As a measure of independence, we consider, for theoretical purpose,
the mutual information:\begin{equation}
  \zZ IY = \int_{\zR^p} P(Y) \log\frac{P(Y)}{\Pi_i P_i(Y_i)} \zdx Y.
\end{equation}
However, as it is too hard to compute, we consider other contrast functions, invariant by permutation, scaling on coordinates and maximal for independent ones.


%% --------- ---- -  -         Information Theory         -  - ---- --------- %%

\subsection{Information Theory}
\label{infth}
Let $X\in\zR^n$ be a random variable, we note $P\zp{X}$ his density and $\Sigma_X$ his covariance matrix.
\znl

In the space of measures, let $\zD G$ be the manifold of Gaussian distributions, $\zD P$ the manifold of ``product'' distributions and $\zD P\wedge\zD G$ the manifold of Gaussian ``product'' distributions. Note that these manifolds are exponential families.

The main advantage of this geometric point of view is that the Kullback-Leibler divergence allows the notion of projection on exponential families. The projection of $P$ on the family $\zD E$ is defined as the vector of $\zD E$ that minimise the divergence to $P$. We write this projection~$P^{\zD E}$.

\znl

Then, we define:

The \textbf{Kullback--Leibler divergence} distribution from $Q$ to $P$:
\begin{equation}
  \Kl PQ = \int_{\zR^n} P(x) \log\frac{P(x)}{Q(x)} \zdx x.
\end{equation}

The \textbf{entropy}:\begin{equation}
  \zZ HP = - \int_{\zR^n} P(x)\log P(x) \zdx x.
\end{equation}

The \textbf{mutual information}:
\begin{equation}
  \zZ IY = \Kl {\zpbig2 P(Y)}{\Pi_i P_i(Y_i)} = \Kl {\zpbig2 P(Y)}{P(Y)^{\zD P}}.
\end{equation}

The \textbf{negentropy}:
\begin{equation}
  \zZ GY = \zZ H{\Ng{\zesp Y}{\Sigma_Y}\zpbig2}-\zZ HY= \zZ H{P(Y)^{\zD G}}-\zZ H{P(Y)}.
\end{equation}


The \textbf{non-gaussianity}:
\begin{equation}
  \zZ GY = \Kl{Y}{\Ng{\zesp Y}{\Sigma_Y}\zpbig2} = \Kl {\zpbig2 P(Y)}{P(Y)^{\zD G}}.
\end{equation}

The \textbf{correlation}:
\begin{equation}\begin{array}{rcl}
  \zZ CY &=& \displaystyle\Kl{\Ng{\zesp Y}{\Sigma_Y}\zpbig2}{\Ng{\zesp Y}{\mathrm{Diag}\:\Sigma_Y}}\\
  &=&\Kl{\zpbig2 P(Y)^{\zD G}}{P(Y)^{\zD P\wedge\zD G}}\\
  &=& \displaystyle\frac 12\log\frac{\det\zp{\mathrm{Diag}(\Sigma_Y)}}{\det\zp{\Sigma_Y}}.
\end{array}\end{equation}

Using the fact that:
\begin{equation}
\Kl {\zpbig2 P(Y)}{P(Y)^{\zD P}}+\sum_i \Kl {\zpbig2 P(Y_i)}{P(Y_i)^{\zD G}} = \Kl {\zpbig2 P(Y)}{P(Y)^{\zD G}} + \Kl{\zpbig2 P(Y)^{\zD G}}{P(Y)^{\zD P\wedge\zD G}}
\end{equation}
We have:
\begin{equation}
        \zZ IY + \sum_i\zZ G{Y_i} = \zZ GY + \zZ CY.
\end{equation}
For more information see~\cite{cardoso2003}.

\subsection{ICA and Maximum Likelihood}
As presented in~\cite{hyvarinen2000}, it is possible to consider ICA as a maximum likelihood problem linked to the infomax principle. 
With the previously introduced notations, the log-likelihood is defined as:
\begin{equation}
L = \sum_t \sum_i \log f_i(w_i^Tx(t))+T\log(|det(W)|)
\end{equation}
Where $f_i$ is the density function of $s_i$. The expectation of this likelihood is:
\begin{equation}
\mathbb{E}[L] = \sum_i \mathbb{E}[\log f_i(w_i^Tx(t))]+\log(|det(W)|)
\end{equation}
In the case where $f_i$ is the actual distribution of $w_i^Tx(t)$, the first term becomes $-\sum_i H(w_i^Tx(t))$ which is one of the independence measures listed in~\ref{infth}.

%% ------- -- -- -  -                                      -  - -- -- --------%%
%---- -- -  -                     Algorithmes                      -  - -- ----%
%%-------- -- -- -  -                                      -  - -- -- ------- %%

\section{Algorithms for ICA}
\subsection{Hérault and Jutten (HJ) algorithm}
This method is based on the neural network principle.
We write $W = \zp{I_n+\widetilde W}^{-1}$ and for a pair of given functions $\zp{f,\:g}$, we adapt $\widetilde W$ as follows:\begin{equation}
\widetilde W_{ij} = f(y_i) g(y_j).
\end{equation}

%% --------- ---- -  -            -----------             -  - ---- --------- %%

\subsection{EASI algorithm}
For a given contrast function $\psi$ and a cost function $J_\psi$, the algorithm iterates as follows:
\begin{equation}
  W_{t+1} = \zp{I_n-\lambda_t\nabla J_\psi(y_t)} W_t.
\end{equation}
The EASI algorithm corresponds to the functions \hbox{$\psi(y) = \sum \zp[b]{y_i}^4$} and $J_\psi(y) = \zesp{\psi(y)}$.

%% --------- ---- -  -            -----------             -  - ---- --------- %%

\subsection{Jade algorithm}
Several methods are based on the cumulants. The goal here is to annul all the cross cumulants of order $4$.
Thus, the idea is to diagonalize the cumulant tensor which is equivalent to minimise the following contrast function:
\begin{equation}
  c\zp{x} = \sum_{i,k,l}\zp[b]{\zop{Cum}{x_i,x_i^*,x_k,x_l}}^2.
\end{equation}

%% --------- ---- -  -            -----------             -  - ---- --------- %%

\subsection{FastICA algorithm}
The FastICA algorithm is based on the information theory. The goal here is to maximise the marginal non-gaussianity on the whitened data, relying on a non linear quadratic function $f$ with the following rule:\begin{equation}
  \widetilde W_{t+1} = \zesp{X.\ztr{f(\ztr{W_t}X)}} - \zesp{f''(\ztr{W_t}X)}W_t,
\end{equation}
with $W_t$ the normalise vector of $\widetilde W_t$. In our experiments, we used $f(x) = \frac{x^4}4$. But it is possible to use $f(x) = \log \cosh x$ or $f(x) = \exp\zp{-\frac{x^2}2}$ as well.



%% ------- -- -- -  -                                      -  - -- -- --------%%
%---- -- -  -                      Resultats                       -  - -- ----%
%%-------- -- -- -  -                                      -  - -- -- ------- %%

\section{Results}
First, we decided to apply the main algorithm to simple simulated data.
For three sinusoidal signals of arbitrary frequencies and phases, we obtain the results in figure \ref{fig:res}, which are good for all methods. We only have to remark that FastICA doesn't retrieve the sign of the signal, which is irrelevant.

%% \begin{figure}\begin{center}
%% \hspace{-.8cm}\begin{mygraph}{xmin=0, xmax=100, %
%%                 ymin=-1, ymax=1, %
%%                 sizex=5, sizey=2.5}%
%%                 {nomx=Sources, nomy=}%
%%                 {0,50,100}{-1,-.5,...,1.05}
%%   \draw[zplot, red] plot file {data/init_sources_2.txt};
%%   \draw[zplot, green!70!black] plot file {data/init_sources_3.txt};
%%   \draw[zplot, blue] plot file {data/init_sources_1.txt};
%% \end{mygraph}\hspace{-1cm}
%% \begin{mygraph}{xmin=0, xmax=100, %
%%                 ymin=-1, ymax=1, %
%%                 sizex=5, sizey=2.5}%
%%                 {nomx=FastICA, nomy=}%
%%                 {0,50,100}{-1,-.5,...,1.05}
%%   \draw[zplot, red] plot file {data/init_fast_1.txt};
%%   \draw[zplot, green!70!black] plot file {data/init_fast_2.txt};
%%   \draw[zplot, blue] plot file {data/init_fast_3.txt};
%% \end{mygraph}

%% \begin{mygraph}{xmin=0, xmax=100, %
%%                 ymin=-1, ymax=1, %
%%                 sizex=5, sizey=2.5}%
%%                 {nomx=HJ, nomy=}%
%%                 {0,50,100}{-1,-.5,...,1.05}
%%   \draw[zplot, red] plot file {data/init_hj_2.txt};
%%   \draw[zplot, green!70!black] plot file {data/init_hj_1.txt};
%%   \draw[zplot, blue] plot file {data/init_hj_3.txt};
%% \end{mygraph}\hspace{-1cm}
%% \begin{mygraph}{xmin=0, xmax=100, %
%%                 ymin=-1, ymax=1, %
%%                 sizex=5, sizey=2.5}%
%%                 {nomx=JADE, nomy=}%
%%                 {0,50,100}{-1,-.5,...,1.05}
%%   \draw[zplot, red] plot file {data/init_jade_3.txt};
%%   \draw[zplot, green!70!black] plot file {data/init_jade_1.txt};
%%   \draw[zplot, blue] plot file {data/init_jade_2.txt};
%% \end{mygraph}
%% \vspace{-.2cm}
%% \caption{Results of the main ICA algorithms on simulated data. Each colour corresponds to a signal. The ``Sources'' graph shows the unmixed signals and the other ones the results of the algorithms. \label{fig:res}}
%% \end{center}\end{figure}


%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                               FIN DU DOCUMENT                                %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%


\bibliographystyle{alpha}
\bibliography{Biblio}{}
\nocite{*}

\label{lastpage}

\end{document}
